---
title: 'IS 605 Homework Week One'
author: "Tom Detzel, Feb. 2, 2018"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# load required packages

suppressMessages(library(easypackages))

suppressMessages(libraries("tidyverse", "pander", "MASS", "stats", "psych", "e1071", "pracma"))

```


## Problem set 1  
  
  
*(1) Calculate the dot product u.v where u = [0.5; 0.5] and v = [3; −4]*

**A.** Also known as the inner product or scalar product, the dot product of two vectors is given by the general formula:

 **v \* u = v^1^, v^2^, ... v^n^ \* u^1^, u^2^, ... u^n^
         = v^1^ \* u^1^ + v^2^ \* u^2^ ... + v^n^ \* u^n^**
         
The result is a scalar.

In R, the operator for a dot product is %*%. Our answer using code is -0.5 as shown below.

```{r}

u <- c(0.5, 0.5)
v <- c(3, -4)
u%*%v

```

*(2) What are the lengths of u and v? Please note that the mathematical notion of the length of a vector is not the same as a computer science definition.*

**A.** The length of a vector is called its *norm* and is computed as the nonnegative square root of the sum of the squares of the vector's components.

The norm (length) of **u** is `r round(abs(sqrt(sum(u^2))),4)`. The norm of **v** is `r abs(sqrt(sum(v^2)))`.

In base R, the length of a vector can be computed using the norm function (choosing the correct type parameter). Examples:

```{r}
norm(u, type="2")
```
```{r}
norm(v, type="2")
```

*(3) What is the linear combination: 3u − 2v?*

**A.** The linear combination is the result of multiplying scalars by each vector and adding the corresponding results.

In this case, we multiply 3 by each element of vector **u** and -2 by each element of vector **v**. We then add the corresponding elements to get a new vector that is the linear combination.

R code follows.

```{r}

linear_combo <- 3*u + (-5*v)
linear_combo

```

*(4) What is the angle between u and v?*  
  
**A.** The angle $\theta$ between two vectors is computed as the dot product of the vectors divided by the product of their lengths. The *cos $\theta$* is a value between -1 and 1.

*cos $\theta$* = **u \* v / ||u||\*||v||**

R code follows:
  
```{r}

angle <- u%*%v/(norm(u, type="2")*norm(v, type="2"))
angle

```
  
## Problem set 2

*Set up a system of equations with 3 variables and 3 constraints and solve for x. Please write a function in R that will take two variables (matrix A & constraint vector b) and solve using elimination.*  
  
### 1 -- Prep the equation set  
  
```{r}

# test matrix given from assignment
#1  1 3    x1 = 1
#2 −1 5    x2 = 2 (1)
#−1 −2 4] [x3] = 6

# create a simple three-variable matrix as given in assignment
A <- matrix(c(1,1,3, 2,-1,5, -1,-2,4), nrow = 3, ncol = 3, byrow = TRUE)
A
# create constraint vector
b <- c(1,2,6)
b

# test the matrix to be sure it's correct
solve(A,b)

```

### 2 -- Write a function to solve 3X3 systems  
  
My solution requires loading the pracma package to compute a row-reduced echelon matrix from the augmented system matrix. I tried to program the RREF code myself but it was taking too long, so I punted on that part. You can see my unfinished scratch work in the appendix. 

Here is my solve function:

```{r}

# function to solve a 3 equation, 3 variable linear system
# assumes input is a 3x3 matrix and 1x3 constraints vector
# requires library pracma

my_solve <- function(A, b){
            # check inputs and quit if wrong
            stopifnot(is.matrix(A), 
                      dim(A)==c(3,3),
                      is.numeric(b),
                      length(b)==3)
            
            # make augmented matrix from inputs
            augA <- as.matrix(cbind(A,b))
            dimnames(augA) <- NULL

            # use pracma pkg to get row-reduced matrix
            # NOTE: I spent hours trying to recreate this and ran out of time
            reduced <- rref(augA)

            # solve for the variables
            x1 <- reduced[1,4] - reduced[1,2] - reduced[1,3]
            x2 <- reduced[2,4] - reduced[2,3] 
            x3 <- reduced[3,4]

            # output the results
            results <- c(x1, x2, x3)
            return(results)
}

# call the function on our system
my_solve(A,b)

```

Try with some other matrices.

```{r}

# from our text, p. 38
A2 <- matrix(c(1,-1,2, 2,1,1, 1,1,0), nrow=3, ncol=3, byrow=T)
b2 <- c(1,8,5)

A3 <- matrix(c(1,2,2, 1,3,3, 2,6,5), nrow=3, ncol=3, byrow=T)
b3 <- c(4,5,6)

my_solve(A2, b2); my_solve(A3, b3)

```
  

  
### Appendix -- Scratch work FWIW
  
Cutting room floor coding from trying to recreate pracma's rref() function.

```{r include=T, eval=F }

# function to compute a row-reduced echelon form matrix using Gauss Jordan elimination and return solution for a three-equation, three-variable linear system

# instantiate variables
j <- 0              # vars count
r <- 0              # pivot rows
m <- dim(augM)[1]   # row count
n <- dim(augM)[2]   # column count

# step 1, order the rows
rowlist <- augM

# first set column index
j <- j+1

# internal function for second row operation to convert i,j to 1
# used to get value for equation operation 2
alpha <- function(x){
 a <- 1/x
 return(a)
}

# test for leading zero, reorder, then do row operation
# later will add for(i in 1:length(m))

if(j == n+1){
  stop("All done.")
} else {
  if(augM[,j]==0){
    j <- j+1
  } else {
    rowlist <- augM[order(augM[,j]),]
    rowlist[1,] <- alpha(rowlist[1,j] * rowlist[1,])
    j <- j +1
  }
}

alpha(rowlist[1,j])

augTest <- augM

# stop if all columns in matrix are checked
if(j == n+1){
  stop("All done.")
} else {
  # if all values in j == 0, move to next column
  if(unique(augTest[,j])==0){
    j <- j+1
  }
} else {
  # select a row in which j is not 0
  # reorder the matrix by row j
  augTest <- augTest[order(augTest[,j], decreasing=T),]
  
  # get index for top row
  i <- augTest[which(augTest[,j]!=0)][1]
  
  # set index for row number
  r <- r+1
  
  # swap the rows
  augTest[c(r,i),] <- augTest[c(i,r),]
  
  # apply second row operation to make i,j==1
  augTest[r,] < augTest[r,] * alpha(augTest[r,j])

}

```

